---
title: "Developing a Workflow to Maximize Reproducibility and Research Impact: Managing Data, Computer Code, and Projects for Success"
author: "Althea A. ArchMiller & John R. Fieberg"
date: "7/12/2017"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Developing a Workflow to Maximize Reproducibility and Research Impact: Managing Data, Computer Code, and Projects for Success

Althea A. ArchMiller &  John R. Fieberg

\centering
\includegraphics[width=2.5in]{images/reproducible_humor_copy.png}
\centering


## Why worry about reproducibility?

\begin{columns}
\begin{column}{0.48\textwidth}
Working towards future reproducibility makes my code easier for my collaborators \\ (and me) to read, run, and debug today,\\ and that’s why I think reproducibility is a \\ \textbf{win-win for all researchers}.”

-Althea
\end{column}
\begin{column}{0.48\textwidth}
\includegraphics[width=2in]{images/phd101212s.png}
\end{column}
\end{columns}


## Why worry about reproducibility?

\centering
\includegraphics[width=3.5in]{images/reproducibility_quote.png}
\centering

\begin{itemize}
\item make your life easier! Now, and in the future
\item collaborations
\item broader research impact
\item increased citations
\item transparency
\item grant and journal requirements
\end{itemize}

“[Reproducibility] provides security, saves time, and forces me to be more thoughtful about my workflow.” - Ethan Young

## Is my research reproducible?

What formats are your research documents stored in? \pause

- .csv
- .txt
- .pdf
- .html
- .R/.Rdata
    - \textcolor{blue}{YES - these are considered "reproducible"} \pause
    
- .doc/.docx
- .sas
- .xls/.xlsx
- any other proprietary file format
    - \textcolor{red}{NO - these are not "reproducible"}

## Is my research reproducible?

- Is your code linear? 
    - Clear environment often and at beginning of script
    - Each program should focus on one main task or analysis
    - Don't rely on manual commenting/uncommenting
    
\pause

```{r, eval=F, echo=T}
# Here, I want to know what variables are significant?
#       I will try each variable in turn
lm.out <- lm(weight ~ height, data = trial.data)
remove(lm.out) # clear previous lm.out for each 
               #   new lm() definition above

# Is the relationship significant? 
#     (If not, clear and try a new variable)
summary(lm.out)
```

        
## Is my research reproducible?

- Are your files easily shared with others?
    - Organized directory structure
    - Files relatively linked
    - Well-documented & commented
    - Consistency in coding practices

"The point of having style guidelines is to have a common vocabulary of coding so people can concentrate on *what* you are saying, rather than on *how* you are saying it." - Google's R Style Guide

## Is my research reproducible?

Do you treat your data as read-only?

- Don't use Excel, etc., to manipulate raw data
- Use an R script for data processing
    - Process data in one script, then save for loading into subsequent scripts
- When archiving, provide raw data and processing code not just final tables

## Workshop Outline

The goal for this workshop is to help you develop the tools to develop a workflow to maximize reproducibility, collaborations, and research impact.

1. RStudio Projects for organizing data, code, and output \pause
2. R-Markdown and R-Oxygen with knitr for documenting your code and creating reproducible reports \pause
3. GitHub for version-control, collaborating and archiving

# 1. RStudio Projects for organizing data, code, and output

## 1: Introduction to RStudio

\includegraphics[width=4.5in]{images/rstudio_annotated.png}

## 1: Introduction to RStudio

\includegraphics[width=4.5in]{images/rstudio_overview.png}

## 1: Introduction to RStudio

\includegraphics[width=4.5in]{images/rstudio_overview2.png}

## 1. RStudio Projects

Think about a typical research project, maybe a dissertation chapter or an experiment that you've managed from data collection through publication. What are typical **folders** that you've used? \pause

- Raw data
- Processed data
- Analysis scripts
- Paper/Manuscript-related documents
- Sharing documents ("transmittals")
- Metadata
- Maps or other deliverables

RStudio Projects provide an opportunity for you to organize and manage all of these types of folders in **one place** in a way that **relatively links** everything together and **eases sharing**.\pause

Up next, Activity 1!    

## Activity 1: Data management and updating

Here, we will read in and process three weeks of experimental data and do some preliminary analysis. Then, we will get a final (4th) week of data, which we will merge with the original data. 

The goals are to:

1. Be introduced to RStudio
2. Create a framework for keeping data organized and up-to-date
3. Automatically update our analyses based on the master dataset

Context: Abundance data from ~75 invertebrate species sampled on various beaches along the Dutch coast. 

*Zuur, A.F., E.N. Ieno, and G.M. Smith (2007) Analysing Ecological Data. Springer, New York.*

## Activity 1: Data management and updating

Before we begin today, we need sync your individual versions of the workshop documents with Althea's master branch:\pause

1. Open RStudio and your reproducibility_workshop.rproj. (File > Open Project...)
2. Open shell (Tools > Shell...)
3. Type in exactly, then press enter:
```{r shell1, eval=F, echo=T}
$ git fetch upstream
```
4. Type in exactly, then press enter:
```{r shell2, eval=F, echo=T}
$ git checkout master
``` 
5.  Type in exactly, then press enter:
```{r shell3, eval=F, echo=T}
$ git merge upstream/master
```

## Activity 1: Data management and updating

1. In the File window of RStudio, copy the **student_folders/student_template** folder. Rename the folder after yourself (or an alias).

2. Open a new R Script file and save it to that new folder as **student_folders/yourname/activity1a_data_processing.R** \pause

Activity Overview:

1. We will read in first three weeks of data and combine them, process the data a little bit, and save the merged/processed data for analysis.

2. We will save another new R Script file as **"activity1b_data_analysis.R"** and code/run some preliminary regression analysis.

3. We will pretend to have just gotten the final week's data in and update everything in a "reproducible" way.
    



# 2. R-Markdown and R-Oxygen with knitr for documenting your code and creating reproducible reports

## Overview of **knitr**

\includegraphics[width=4.5in]{images/knitr_overview.png}

## Why **knitr** for manuscripts?


\centering
\includegraphics[width=3in]{images/knitr_quote.png}
\centering 

\pause

"I can do reproducible work in R (making me happy) and format the output report in Word (making my collaborators happy)" - Richard Layton http://rmarkdown.rstudio.com/articles_docx.html

## Why **knitr** for manuscripts?

Native R Scripts (.R extensions) (or any analysis code) are generally not designed for reading, but the **knitr** library has been designed for converting R scripts into readable reports, such as Word, PDF, and/or html documents.

Not only do these types of reports help with collaborating, they provide a great framework for archiving your analyses and results.

Example: https://conservancy.umn.edu/handle/11299/181607

## Documenting Code: General tips

- Consistent and meaningful naming conventions
    + \textcolor{red}{a = b*c}
    + weekly.pay = hours.worked*pay.rate (not cross-compatible)
    + weekly_pay = hours_worked*pay_rate
    + weeklyPay = hoursWorked*payRate \pause
- Use YYYYMMDD or equivalent for dates

\includegraphics[width=3in]{images/dissertation_data.png}

## Documenting Code: R-Markdown

R-Markdown combines \textcolor{blue}{markdown} language, which is "an easy-to-write plain text format" and embedded \textcolor{blue}{R code chunks} that are "run so their output can be included in the final document" [1] \pause

\begin{columns}
\begin{column}{0.48\textwidth}
\includegraphics[width=2in]{images/rmd1.png}
\end{column}
\begin{column}{0.48\textwidth}
\includegraphics[width=2in]{images/rmd2.png}
\end{column}
\end{columns}

[1] www.rmarkdown.rstudio.com

## Documenting code: R-Markdown

Exercise 2a: Introduction to R-Markdown

- File > New File > R Markdown...
- Choose "html" - optionally put in a title and press "OK"
- This R-Markdown template is ready to "knit" into an html as-is
    + Click the blue Knit button 
    + Save as **"student_folders/yourname/activity2a_intro_rmarkdown.Rmd"**
    + View the resultant html
- Take a few minutes to modify the .Rmd and view how the changes appear in the knit html document.
    
<https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf>

## Documenting code: ezknitr

What folder did the html end up in? \pause

Now imagine if you wanted to keep the programs/scripts in a folder separate from reports (\textcolor{blue}{highly recommended!}). You can easily direct the output html file into a different folder using **ezknitr** package. \pause

\small
```{r ezknitr, eval=FALSE, echo=T}
library(ezknitr)
ezknit("student_folders/yournm/activity2a_intro_rmarkdown.Rmd",
       out_dir = "student_folders/yournm/reports", 
       fig_dir = "figures",
       keep_md = F)
```

*Note: When using ezknit, must manually save the .Rmd file before knitting!*

## Documenting Code: R-Oxygen

Instead of using the R-Markdown language, you can also use **pure R scripts** plus **Roxygen comments** (#') to create fully reproducible reports. \pause

Benefit: The entire program can be written and run in the familiar R Script file, then "spun" into an html/Word/pdf document at any point.

- Learning to code with R-Oxygen is arguably more natural since we already use # for commenting \pause
    
Additionally, use #+ to define and label R chunks like we did with the ```{r ...} code in the R-Markdown language.

## Documenting Code: R-Oxygen

Activity 2b: Introduction to R-Oxygen

Here, we'll quickly convert "activity1b_data_analysis.R" into an html document.

1. Open "activity1b_data_analysis.R" 
2. Save As... "student_folder/yourname/activity2b_intro_roxygen.R"

## LaTeX

Another benefit of using knitr/Rmd/Roxygen for creating statistical reports is the nice interface with LaTeX equation syntax.

**Activity 3. LaTeX Equations**

1. Create a new .Rmd html document
2. Save it as "student_folders/yourname/activity3_latex.Rmd"
3. Create the following in the output html
    + $\alpha + \beta = 2\theta$
    + $\pi^2 = 9.86$
    + $\sum_{i=1}^n \sqrt{i} = 42$ (advanced)
4. When you "knit" remember to use ezknitr:

\small
```{r ezknitr2, eval=FALSE, echo=T}
library(ezknitr)
ezknit("student_folders/yourname/activity3_latex.Rmd",
       out_dir = "student_folders/yourname/reports", 
       fig_dir = "figures",
       keep_md = F)
```

<https://tobi.oetiker.ch/lshort/lshort.pdf> (hint: tables on p75)

## Project Workflow w/ RStudio & knitr

Example project directory

- data/
    + raw_data/
    + processed_data/
    + output_data/
- manuscript/
    + ms_figures/
    + transmissions/
    + submission/
- output/
    + figures/
- programs/
- project_file.Rproj

## Project Workflow w/ RStudio & knitr

- **data/**
    + **raw_data/**
        + survey_data20161227.csv
        + survey_data20161230.csv
        + survey_data20170103.csv
    + **processed_data/**
        + survey_data_all.Rdata
    + **output_data/**
        + model_out.Rdata 
        
## Project Workflow w/ RStudio & knitr

- **programs/**
    + a_data_processing.R
    + b_data_analysis.R
    + c_plots.R 
    
## Project Workflow w/ RStudio & knitr

- **output/**
    + a_data_processing.html
    + b_data_analysis.html
    + c_plots.html
    + **figures/**
        + eda1.jpg
        + scatter1.jpg
        
## Project Workflow w/ RStudio & knitr

- **manuscript/**
    + ms.Rmd
    + ms.pdf
    + ms.docx
    + **ms_figures/**
        + fig1.jpg
        + fig2.jpg\pause
    + **transmittals/** 
        + **from_john/**
        +   ms20170523.docx
        +   ms20170625.docx
        + **from_bob/**
        +   ms20170626.docx \pause
    + **submission/** 
        + ms.pdf
        + fig1.pdf
        + fig2.pdf
        + coverletter.docx
        
## Project Workflow w/ RStudio & knitr

Example of an RStudio project that Althea & John used from conceptualization through publication. \pause

### Activity 3: Creating a reproducible report using R-Markdown

Tasks:

1. Knit the rmd as is into a .doc file
2. Add a third week of data and update report
3. Change the formatting using the word-styles-reference-01.docx
        
# 3. GitHub for version-control, collaborating and archiving

## Introduction to GitHub

GitHub provides a place for you to back-up and version-control your R projects.

- Interfaces directly with RStudio
- Free (public repositories or with university email) or cheap (private repositories)
- Facilitates collaboration with other co-authors (or the public)
- Can go back to previous code versions (version-control aspect)

Once you're up and running, it's (usually) simple to use!

## Introduction to GitHub

\centering
\includegraphics[width=3.3in]{images/github_quote2.png}
\centering
\pause

\centering
\includegraphics[width=3.3in]{images/github_quote1.png}
\centering
        
## GitHub Overview

\includegraphics[width=4.5in]{images/github_overview_presi.png}
        
## Activity 4: GitHub in RStudio

First, we will use the built-in interface for GitHub within RStudio. 

Secondly, we will use the most common git commands in shell. \pause

```{r shell, eval=F,echo=T}
git pull origin master # To pull
git add data # add files to a commit
git commit -m "commit message here" #commit
git push origin master # To push commit(s)
```

        
Additional GitHub Tips

- Don't use github with large files (push limit of 100MB, warning >50MB)
    - If necessary, use git-large-file add on (advanced!)
- Create new projects in GitHub first, then sync them with RStudio (File > New Project > Version Control > Git)

## Thanks!

We can use your feedback for the workshop at TWS. Please take a moment or two to fill out our survey.

<https://docs.google.com/forms/d/e/1FAIpQLSdmvePcbb3wztx2JpJUhgTqpjvbR0z3etpnjGrRSRfTBnLpcQ/viewform?usp=sf_link>

\centering
\includegraphics[width=3.5in]{images/git_commit.png}
\centering

## Additional resources

- <https://swcarpentry.github.io/r-novice-gapminder/02-project-intro/>

- knitr documentation and help
<https://yihui.name/knitr/>

- Markdown basics 
<https://daringfireball.net/projects/markdown/basics>

- R-Oxygen formatting/tips <https://rpubs.com/alobo/spintutorial>

- Online Reproducible Research Course
<http://eriqande.github.io/rep-res-web/>
