---
title: "Developing a Workflow to Maximize Reproducibility and Research Impact: Managing Data, Computer Code, and Projects for Success"
author: "Althea A. ArchMiller & John R. Fieberg"
date: "7/12/2017"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome!

### Developing a Workflow to Maximize Reproducibility and Research Impact: Managing Data, Computer Code, and Projects for Success

Althea A. ArchMiller *Assistant Professor, Biology, Concordia College, MN*

John R. Fieberg *Associate Professor of Quantitative Ecology, University of Minnesota*

\textcolor{blue}{Before we begin, please open up the reproducible\_workshop repository in RStudio and install these packages}:

- knitr
- ezknitr
- devtools

## Why worry about reproducibility?

\begin{columns}
\begin{column}{0.48\textwidth}
Working towards future reproducibility makes my code easier for my collaborators \\ (and me) to read, run, and debug today,\\ and that’s why I think reproducibility is a \\ \textbf{win-win for all researchers}.”

-Althea
\end{column}
\begin{column}{0.48\textwidth}
\includegraphics[width=2in]{images/phd101212s.png}
\end{column}
\end{columns}


## Why worry about reproducibility?

“[Reproducibility] provides security, saves time, and forces me to be more thoughtful about my workflow.” - Ethan Young

\begin{itemize}
\item make your life easier! Now, and in the future
\item collaborations
\item broader research impact
\item increased citations
\item transparency
\item grant and journal requirements
\end{itemize}

## Is my research reproducible?

- Are your research documents stored in these formats?
    - .csv
    - .txt
    - .pdf
    - .html
    - .R/.Rdata
        - \textcolor{blue}{YES!}
        
    - .doc/.docx
    - .sas
    - .xls/.xlsx
    - any other proprietary file format
        - \textcolor{red}{NO!}

## Is my research reproducible?

- Is your code linear? 
    - Clear environment often and at beginning of script
    - Don't save .Rdata or history
    - Each program should focus on one main task or analysis
    - Don't rely on manual commenting/uncommenting \pause

So, what's wrong here?

```{r, eval=F, echo=T}
# What variables are significant?
lm.out <- lm(weight ~ height, data = trial.data)
remove(lm.out) # clear previous lm.out for each 
               #   new lm() definition above

# Is the relationship significant? 
#     (If not, clear and try a new regressor)
summary(lm.out)
```

        
## Is my research reproducible?

- Are your files easily shared with others?
    - Organized directory structure
    - Files relatively linked
    - Well-documented & commented
    - Consistency in coding practices

"The point of having style guidelines is to have a common vocabulary of coding so people can concentrate on *what* you are saying, rather than on *how* you are saying it." - Google's R Style Guide

## Workshop Outline

The goal for this workshop is to help you develop the tools to develop a workflow to maximize reproducibility, collaborations, and research impact.

1. RStudio Projects for organizing data, code, and output \pause
2. R-Markdown and R-Oxygen for documenting your code and creating reproducible reports \pause
3. GitHub for version-control, collaborating and archiving

## 1. RStudio Projects

Think about a typical research project, maybe a dissertation chapter or an experiment that you've managed from data collection through publication. What are typical **folders** that you've used? \pause

- Raw data
- Processed data
- Analysis scripts
- Paper/Manuscript-related documents
- Sharing documents ("transmittals")
- Metadata
- Maps or other deliverables

RStudio Projects provide an opportunity for you to organize and manage all of these types of folders in **one place** in a way that **relatively links** everything together and **eases sharing**.\pause

Up next, Activity 1!    

## Activity 1: Data management and updating

Here, we will read in and process three weeks of experimental data and do some preliminary analysis. Then, we will get a final (4th) week of data, which we will merge with the original data. 

The goals are to:

1. Be introduced to RStudio
2. Create a framework for keeping data organized and up-to-date
3. Automatically update our analyses based on the master dataset

Context: Abundance data from ~75 invertebrate species sampled on various beaches along the Dutch coast. 

*Zuur, A.F., E.N. Ieno, and G.M. Smith (2007) Analysing Ecological Data. Springer, New York.*

## Activity 1: Data management and updating

Before we begin today, we need sync your individual versions of the workshop documents with Althea's master branch:\pause

1. Open RStudio and your reproducibility_workshop.rproj. (File > Open Project...)
2. Open shell (Tools > Shell...)
3. Type in exactly, then press enter:
```{r shell1, eval=F, echo=T}
$ git fetch upstream
```
4. Type in exactly, then press enter:
```{r shell2, eval=F, echo=T}
$ git checkout master
``` 
5.  Type in exactly, then press enter:
```{r shell3, eval=F, echo=T}
$ git merge upstream/master
```

## Activity 1: Data management and updating

Now create a new folder in student_folders/ for all of today's activities. Name the folder after yourself (or an alias).

Open a new R Script file and save it to that new folder as **"activity1a_data_processing.R"**

First, we will read in first three weeks of data and combine them, process the data a little bit, and save the merged/processed data for analysis.

Secondly, we will save another new R Script file as **"activity1b_data_analysis.R"** and do (preliminary) regression analysis.

Finally, we will pretend to have just gotten the final week's data in and update everything in a "reproducible" way.
    
## 1. RStudio Projects

**Other links**

https://swcarpentry.github.io/r-novice-gapminder/02-project-intro/

**Data Mangement Tips**

- Treat data as read-only
    - Don't use Excel, etc, to manipulate raw data
    - Use a single R program for all manipulation
    - Save "cleaned" or "processed" data in easily loadable formats 
- Differentiate data types with folders *raw* versus *processed* versus *output* (e.g., linear regression objects, etc)
- Write dates in YYYYMMDD or equivalent format

## Tips

- Don't use github with large files :-(
- Create new projects in GitHub first, then sync them with RStudio


## Why R-Markdown for manuscripts?

"I can do reproducible work in R (making me happy) and format the output report in Word (making my collaborators happy)" - Richard Layton http://rmarkdown.rstudio.com/articles_docx.html

## Documenting Code

Native R Scripts (.R extensions) (or any analysis code) are generally not designed for reading, but the **knitr** library has been designed for converting R scripts into readable reports, such as Word, PDF, and/or html documents.

Not only do these types of reports help with collaborating, they provide a great framework for archiving your analyses and results.

Example: https://conservancy.umn.edu/handle/11299/181607

## Documenting Code

General tips for better coding:

- Consistent and meaningful naming conventions
    + \textcolor{red}{a = b*c}
    + weekly.pay = hours.worked*pay.rate (not cross-compatible)
    + weekly_pay = hours_worked*pay_rate
    + weeklyPay = hoursWorked*payRate
- Using YYYYMMDD or equivalent for dates
- When possible, bundle data, code, figures/tables together with **knitr** package using R-Markdown or R-Oxygen languages

## Documenting Code: R-Markdown

R-Markdown combines \textcolor{blue}{markdown} language, which is "an easy-to-write plain text format" and embedded \textcolor{blue}{R code chunks} that are "run so their output can be included in the final document" [1] \pause

\begin{columns}
\begin{column}{0.48\textwidth}
\includegraphics[width=2in]{images/rmd1.png}
\end{column}
\begin{column}{0.48\textwidth}
\includegraphics[width=2in]{images/rmd2.png}
\end{column}
\end{columns}

[1] www.rmarkdown.rstudio.com

## Documenting code: R-Markdown

Exercise 2a: Introduction to R-Markdown

- File > New File > R Markdown...
- Choose "html" - optionally put in a title and press "OK"
- This R-Markdown template is ready to "knit" into an html as-is
    + Click the blue Knit button 
    + Save in your student_project folder as "activity2a_intro_rmarkdown.Rmd"
    + See the resultant html
- Take a few minutes to modify the .Rmd and view how the changes appear in the knit html document.
    
<https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf>

## Documenting code: ezknitr

What folder did the html end up in? \pause

Now imagine if you wanted to keep the programs/scripts in a folder separate from reports (\textcolor{blue}{highly recommended!})? You can easily direct the output html file into a different folder using **ezknitr** package.

\small
```{r ezknitr, eval=FALSE, echo=T}
library(ezknitr)
ezknit("student_folders/yours/activity2a_intro_rmarkdown.Rmd",
       out_dir = "student_folders/archmiller_a/reports", 
       fig_dir = "figures",
       keep_md = F)
```

## Documenting Code: R-Oxygen

Instead of using the R-Markdown language, you can also use **pure R scripts** plus **Roxygen comments** (#') to create fully reproducible reports! \pause

Benefit: The entire program can be written and run in the familiar R Script file, then "spun" into an html/Word/pdf document at any point.
    + Learning to code with R-Oxygen is arguably more natural since we already use # for commenting \pause
    
Additionally, use #+ to define and label R chunks like we did with the ```{r ...} code in the R-Markdown language. \pause

## Documenting Code: R-Oxygen

Activity 2b: Introduction to R-Oxygen

Here, we'll quickly convert "activity1b_data_analysis.R" into an html document!

1. Open "activity1b_data_analysis.R" 
2. Save As... "student_folder/yours/activity2b_intro_roxygen.R"

## LaTeX

(Page 75 begins the mathematical symbols)

https://tobi.oetiker.ch/lshort/lshort.pdf