---
title: "Developing a Workflow to Maximize Reproducibility and Research Impact: Managing Data, Computer Code, and Projects for Success"
author: "John R. Fieberg & Althea A. ArchMiller"
date: "4/11/2017"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Why worry about reproducibility?

\begin{columns}
\begin{column}{0.48\textwidth}
Working towards future reproducibility makes my code easier for my collaborators \\ (and me) to read, run, and debug today,\\ and that’s why I think reproducibility is a \\ \textbf{win-win for all researchers}.”

-Althea
\end{column}
\begin{column}{0.48\textwidth}
\includegraphics[width=2in]{images/phd101212s.png}
\end{column}
\end{columns}


## Why worry about reproducibility?

“[Reproducibility] provides security, saves time, and forces me to be more thoughtful about my workflow.” - Ethan Young

\begin{itemize}
\item make your life easier!
\item collaborations
\item broader research impact
\item increased citations
\item transparency
\item grant and journal requirements
\end{itemize}

## Is my research reproducible?

- Are your research documents stored in these formats?
    - .csv
    - .txt
    - .pdf
    - .html
    - .R
        - \textcolor{blue}{YES!}
        
    - .doc/.docx
    - .sas
    - .xls/.xlsx
    - any other proprietary file format
        - \textcolor{red}{NO!}

## Is my research reproducible?

- Is your code linear? 
    - Clear environment often and at beginning of script
    - Don't save .Rdata or history
    - Each program should focus on one main task or analysis
    - Don't rely on manual commenting/uncommenting

```{r, eval=F, echo=T}
# What variables are significant?
lm.out <- lm(weight ~ height, data = trial.data)
remove(lm.out) # clear previous lm.out for each 
               #   new lm() definition above

# Is the relationship significant? 
#     (If not, clear and try a new regressor)
summary(lm.out)
```

        
## Is my research reproducible?

- Are your files easily shared with others?
    - Organized directory structure
    - Files relatively linked
    - Well-documented & commented
    - Consistency in coding practices

"The point of having style guidelines is to have a common vocabulary of coding so people can concentrate on *what* you are saying, rather than on *how* you are saying it." - Google's R Style Guide

## Workshop Outline

The goal for this workshop is to help you develop the tools to develop a workflow to maximize reproducibility, collaborations, and research impact.

1. RStudio Projects for organizing data, code, and output
2. R-Markdown and R-Oxygen for documenting your code
3. GitHub for version-control, collaborating and archiving

## 1. RStudio Projects

Think about a typical data analysis project, maybe a dissertation chapter or an experiment that you've managed from data collection through publication. What are typical **folders** that you've used?

- Raw data
- Processed data
- Analysis scripts
- Paper/Manuscript-related documents
- Sharing documents ("transmittals")
- Metadata
- Maps or other deliverables

RStudio Projects provide an opportunity for you to organize and manage all of these types of folders in **one place** in a way that **relatively links** everything together and **eases sharing**.

## 1. RStudio Projects

**Tips**

- Treat data as read-only
    - Don't use Excel, etc, to manipulate raw data
    - Use a single R program for all manipulation
    - Save "cleaned" or "procesed" data in easily loadable format
    
## 1. RStudio Projects

**Other links**

https://swcarpentry.github.io/r-novice-gapminder/02-project-intro/



## Tips

- Don't use github with large files :-(
- Create new projects in GitHub first, then sync them with RStudio


## Why R-Markdown for manuscripts?

"I can do reproducible work in R (making me happy) and format the output report in Word (making my collaborators happy)" - Richard Layton http://rmarkdown.rstudio.com/articles_docx.html

